---
title: "USP 634 Lab7- ANOVA"
author: "Jamaal Green"
date: "May 10, 2016"
output:
  html_document: default
  pdf_document: default
---
```{r, message=FALSE, warning=FALSE}

#load up our required packages for the lab
if (!require("printr")) install.packages("printr", repos='http://yihui.name/xran')
if(!require("car")) install.packages("car", repos = 'http://cran.us.r-project.org')
if(!require("ggplot2")) install.packages("ggplot2", repos = 'http://cran.us.r-project.org')
if(!require("dplyr")) install.packages("dplyr", repos = 'http://cran.us.r-project.org')
if (!require("pacman")) install.packages("pacman")
p_load(printr, car, ggplot2, dplyr)
```

**Learning Objectives**
* Run ANOVA

Download and load the pums dataset ("eb_pums_2000.rds") used previously. Apply the recodes we did in lab 6 to the dataset using the syntax below. 

```{r, warning=FALSE, message=FALSE}
#remember to set your working director
pums <- readRDS("eb_pums_2000.rds")

```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
pums <- pums %>% select(year, metaread, puma, ownershp, builtyr, hhincome, perwt, age, sex, race, hispan, tranwork, carpool, trantime)

#Rename a variable: metaread to detailed.meta.area
names(pums)[2] <- 'detailed.meta.area'

# the code book for pums data can be found at #https://usa.ipums.org/usa-action/variables/group?id=race_eth
pums$racehisp <- NA  
pums$racehisp[pums$race =='White'] <- 1 

#Recode African-American. Enter a value of 2 for those cases where race is African-American: 
pums$racehisp[as.integer(pums$race) ==2] <- 2  

#Recode Asian.  To recode the 3 Asian categories to 3 enter (the "|" sign indicates "or"): 
pums$racehisp[as.integer(pums$race) ==4 | as.integer(pums$race) ==5 | as.integer(pums$race) ==6] <- 3 

#OR, use the following command:
#pums$racehisp[as.integer(pums$race) %in% c(3,4,5)] <- 3

#Recode Hispanic.  Code all Hispanic subcategories as "4" ("!=" means "does not equal"): 
pums$racehisp[as.integer(pums$hispan)!=1] <- 4 

#Note: People of any race can also be Hispanic (Hispanic is not a race category), 
#so when you recode Hispanics as "4", they will no longer be coded in the "White," "Black" or "Asian" categories. 

#Recode Other Race.  Create the residual "Other" category: 
pums$racehisp[is.na(pums$racehisp)] <- 5


#Check your work on the first 30 rows of dataset. A faster way to check your data
pums[1:30,c('race','hispan', 'racehisp')]



#Look at Your New Data.
table(pums$racehisp)
xtabs(~race + hispan, data = pums)

#Label your new race categories with text labels.
#Use the following command to apply text labels to your new race categories: 
pums$racehisp <- factor(pums$racehisp, levels=1:5, 
                       labels=c('White non-Hisp', 'Black non-Hisp', 'Asian non-Hisp', 'Hispanic', 'Other'))

#Look at your work. 
table(pums$racehisp)


table(pums$tranwork)

pums$tranwork <- factor(pums$tranwork, 
                        levels=c(00, 10, 11, 12, 13, 14, 15, 20, 30, 31, 32, 33, 34, 35, 36, 40, 50, 60, 70),
                        labels=c("N/A (+ not reported 1960)", "Auto, truck, or van", "Auto", "Driver", "Passenger", "Truck", "Van", "Motorcycle", "Bus or streetcar", "Bus or trolley bus", "Streetcar or trolley car", "Subway or elevated", "Railroad", "Taxicab", "Ferryboat", "Bicycle", "Walked only", "Other", "Worked at home"))

# recode modes
pums$mode = as.integer(pums$carpool)  # this creates a copy of "carpool" with drive-alone & carpool commuters
pums$mode[as.integer(pums$tranwork) %in% c(10, 11, 12, 13, 14, 15)] <- 3  # this creates a "transit" category
pums$mode[as.integer(pums$tranwork) %in% c(16, 17)] <- 4  # this creates "bike/walk"
pums$mode[as.integer(pums$tranwork) %in% c(8, 18, 19)] <- 5 # this creates "other"
pums$mode[pums$mode==0] <- NA  #this sets NA where the original "carpool" variable to #missing
pums$mode <- factor(pums$mode, levels=1:5, labels=c('Drive alone', 'Carpool', 'Transit', 'Bike/walk', 'Other')) 

comment(pums$mode) <- "Mode of transportation to work (recoded from carpool and tranwork)"

# recode builtyr variable
pums$builtyr2 <- as.integer(pums$builtyr)  # copy of builtyr
pums$builtyr2[as.integer(pums$builtyr) %in% c(2, 3)] <- 1 #recodes first two categories to create 0-10 years
pums$builtyr2[as.integer(pums$builtyr)==4] <- 2 #set numeric codes so they are equivalent to # decades
pums$builtyr2[as.integer(pums$builtyr)==5] <- 3
pums$builtyr2[as.integer(pums$builtyr)==6] <- 4
pums$builtyr2[as.integer(pums$builtyr)==7] <- 5
pums$builtyr2[as.integer(pums$builtyr)==8] <- 6
pums$builtyr2[as.integer(pums$builtyr)==9] <- 7

pums$builtyr2 <- factor(pums$builtyr2, levels=1:7, labels=c("0-10 yrs", "11-20 yrs", "21-30 yrs", "31-40 yrs", "41-50 yrs","51-60 yrs", "61+"))

comment(pums$builtyr2) <- "recode of builtyr, categories equal one decade"

pums$increc <- pums$hhincome
pums$increc[pums$hhincome <=0 | pums$hhincome >= 999999] <-  NA

pums$tenure = as.integer(pums$ownershp)
pums$tenure[pums$ownershp==0] <- NA
pums$tenure[pums$ownershp==1] <- "Owned"
pums$tenure[pums$ownershp==2] <- "Rented"

table(pums$tenure)

# keep 2 decimal places in output
options(digits=2)

#save data frames
#save(data, pums, file="eb_pums_2000_v2.RData")

#View Frequency Distribution
table(pums$mode)
table(as.integer(pums$mode))

#Cross-tabulate means of travel to work by race/ethnicity
xtabs(~mode+racehisp,data=pums,subset=!is.na(mode), drop.unused.levels=T)

#assign cross-tab to a variable
xtabs.TR = xtabs(~mode+racehisp, data=pums, subset=!is.na(mode), drop.unused.levels=T)
xtabs.TR
#Get column %
prop.table(xtabs(~mode+racehisp,data=pums,subset=!is.na(mode), drop.unused.levels=T), 2)
#OR 
prop.table(xtabs.TR,2)

#results of chi-square test of independence
summary(xtabs.TR)
#alternatively


```



##Boxplot by group

We will create a boxplot of trantime by racehisp using the *ggplot2* package.

```{r, message=FALSE, warning=FALSE}

p1 <- ggplot(pums, aes(x = mode, y = trantime))
p1 + geom_boxplot()

```

##ANOVA

Now we will conduct an ANOVA test using the PUMS dataset in order to determine whether commuting travel time (trantime) varies significantly by race/Hispanic origin- i.e. we will test the likielihood that the sample data come from a population in which personal income does not vary between racial/ethnic categories. 

**Conduct ANOVA test**
Recall that for ANOVA to work, we need the groups to be of equal size and equal variance. We can check the group size with a bar chart and the variances by group.

```{r, warning=FALSE, message=FALSE}
#for our frequency bar chart
p2 <- ggplot(pums, aes(x = racehisp))
p2 + geom_bar()

#for our variance by group
sdbygroup <- pums %>% group_by(racehisp) %>%
  summarize(std_dev = sd(trantime))

sdbygroup
```

Since non-Hispanic Whites outnumber many of the other racial/ethnic groups we will limit the test to only three groups: Black non-Hispanic, Asian non-Hispanic, and Hispanic. Let's create a subset for the three groups.

```{r, message=FALSE, warning=FALSE}

#base r approach
pums.sub <-pums[pums$racehisp %in% c("Hispanic", "Black non-Hisp", "Asian non-Hisp"),]
#check
table(pums.sub$racehisp)
#dplyr approach
pums.sub2 <- pums %>% filter(racehisp == "Hispanic" | racehisp == "Black non-Hisp" | racehisp == "Asian non-Hisp")

#check
table(pums.sub2$racehisp)

#Recompare the sample sizes
p3 <- ggplot(pums.sub2, aes(x = racehisp))
p3 + geom_bar()
```

Now that we have similar sample sizes we have met the assumptions for conducting an ANOVA. (Note that you can also draw random samples from your existing sample in order to produce groups of equal size)

What are the other critical assumptions of ANOVA again?

As with all things R there are many ways of conducting an ANOVA. We will use the *aov* function as one approach. The formula for ANOVA follows the general form of:

> aov(dependent variable ~ grouping variable, data = df)

The dependent variable is on the left side of ~ while the independent variable (the variable we are looking to test whether or not there is a difference between its groups) is on the other side. 

```{r, warning=FALSE, message=FALSE}

mod1 <- aov(trantime ~ racehisp, data = pums.sub)

summary(mod1)
```

Examine the outputs in the ANOVA table, and determine whether to reject the null hypothesis. Zero in on the F-statistics, and its probability value (and significance codes); this indicates the probability that the null hypothesis can be rejected.

*Pairwise comparison of means*

For the pairwise comparison of means we will use the *pairwise.t.test()* function. Are the differences from other groups significant? What is the adjustment using the Bonferroni correction for multiple comparisons. The general form of the pairwise test function is this:

> pairwise.t.test(dependent var, independent var,
p.adj = c("none", "bonferroni", "holm"))

```{r, message=FALSE, warning=FALSE}
pairwise.t.test(pums.sub$trantime , pums.sub$racehisp, p.adj = "bonferroni")

```

Quick question: Conduct an ANOVA test in order to determine whether personal income (increc) varies significantly by race/Hispanic origin. 

##Test of Equal Variances
We can use hypothesis testing to test whether groups have equal variance. One such test is Levene's Equality of Variance Test. Levene's test estimates a hypothesis test on whether the variances of your groups are equal. Thus the the null hypothesis is that the variances are in fact equal. What does this mean for our alternative hypothesis?

```{r, message=FALSE, warning=FALSE}

#running Leven's Equality of Variance
leveneTest(trantime ~ racehisp, data = pums.sub,
  center = "mean")
```