---
title: 'USP 634 Lab 6: Chi-Square!'
author: "Jamaal Green"
date: "May 3, 2016"
output:
  html_document: default
  word_document: default
---

##Learning Objectives
1. Recode variables; working with factors
2. Run crosstabs
3. Run Chi-Square Test

First, either create a new folder for your working directory or use a directory you've already made. Download the eb_pums dataset for the lab.
Load the eb_pums dataset using the **readRDS()** function.

```{r, warning=FALSE, message=FALSE}

pums <- readRDS("eb_pums_2000.rds")

```


###Recoding Variables

We will recode some of the values in pums dataset so that we can run our tests later on. 
```{r, echo=TRUE, message=FALSE, warning=FALSE}

#getting recodes for pums df. racehisp variable and the mode variable
pums$racehisp <- NA  
pums$racehisp[pums$race =='White'] <- 1 

pums$racehisp[as.integer(pums$race) ==2] <- 2  

#Recode Asian.  To recode the 3 Asian categories to ???3???, enter (the ???|??? sign indicates ???or???): 
pums$racehisp[as.integer(pums$race) ==4 | as.integer(pums$race) ==5 | as.integer(pums$race) ==6] <- 3 

pums$racehisp[as.integer(pums$hispan)!=1] <- 4 

#Note: People of any race can also be Hispanic (Hispanic is not a race category), 
#so when you recode Hispanics as ???4???, they will no longer be coded in the ???White,??? ???Black,??? or ???Asian??? categories. 

#Recode Other Race.  Create the residual ???Other??? category: 
pums$racehisp[is.na(pums$racehisp)] <- 5

pums$racehisp = factor(pums$racehisp, levels=1:5, 
                       labels=c('White non-Hisp', 'Black non-Hisp', 'Asian non-Hisp', 'Hispanic', 'Other'))

pums$tranwork <- factor(pums$tranwork, 
                        levels=c(00, 10, 11, 12, 13, 14, 15, 20, 30, 31, 32, 33, 34, 35, 36, 40, 50, 60, 70),
                        labels=c("N/A (+ not reported 1960)", "Auto, truck, or van", "Auto", "Driver", "Passenger", "Truck", "Van", "Motorcycle", "Bus or streetcar", "Bus or trolley bus", "Streetcar or trolley car", "Subway or elevated", "Railroad", "Taxicab", "Ferryboat", "Bicycle", "Walked only", "Other", "Worked at home"))

# recode modes
pums$mode <- as.integer(pums$carpool)  # this creates a copy of "carpool" with drive-alone & carpool commuters
pums$mode[as.integer(pums$tranwork) %in% c(10, 11, 12, 13, 14, 15)] <- 3  # this creates a "transit" category
pums$mode[as.integer(pums$tranwork) %in% c(16, 17)] <- 4  # this creates "bike/walk"
pums$mode[as.integer(pums$tranwork) %in% c(8, 18, 19)] <- 5 # this creates "other"
pums$mode[pums$mode==0] = NA  #this sets NA where the original "carpool" variable to missing
pums$mode = factor(pums$mode, levels=1:5, labels=c('Drive alone', 'Carpool', 'Transit', 'Bike/walk', 'Other')) 
comment(pums$mode) = "Mode of transportation to work (recoded from carpool and tranwork)"


```
Download the Lab6_recodevar.R script and open it. It should automatically open in your R-Studio console. Next, we will rename our "data" dataframe to a slighty more expressive name "pums".



**Rename a variable in a dataframe**
Find the location of the variable you want to rename-- in this case, change the 6th variable "metaread" to "detailed.meta.area".

```{r, warning=FALSE, message=FALSE}

names(pums)[6] <- "detailed.meta.area"
```

Check your table to ensure the data have been recoded correctly.

Renaming and recoding variables allow us to solve some of the problems we encountered in the first assignment where we may have too many variables.

**Note** You have to code your categorical variables as *factors* before R can recognize them as true categorical variables and not simply text. 

##Run Crosstablulations

**Crosstabulate means of travel to work by race/ethnicity**

Cross tabs are one of the most basic and commonly used means to analyze categorical variables. We will crosstab two categorical variables and produce an analytical table using percentages.
 1. First let's simply get a table of mode. We will use the **table()** function.
 
 
```{r, warning=FALSE, message=FALSE}
 table(pums$mode)
 table(as.integer(pums$mode))
```

Quick question: What numerical code designates an NA? What does it mean?
Quicker answer: The "universe" for this variable is all workers, and NA refers to all non-workers.

Now we shall add our second categorical variable and finish our crosstabulation. We will crosstabe "mode" with "racehisp" (your recoded Race/Hispanic variable). We shall make the race/ethnicity categories our column headers and travel model our row labels. The general mode for getting a crosstabs is of the form:

> xtabs(~var1 + var2, data = df)




*Limit our crosstabs to workers only*

We can use the **subset()** function to isolate particular values of our variables. In this case, we are only interested in workers. (we will save the table object as xtabs.TR)

```{r, warning=FALSE, message=FALSE}

xtabs.TR <- xtabs(~ racehisp + mode, data = pums, 
                  subset = !is.na(mode), drop.unused.levels = TRUE)

xtabs.TR
```

*Add column percentages*

We have a nice proper crosstab but frequency counts are often not what we want. We are often interested in the different *proportions* of our categorical variables. So, to answer this we will calculate the percentages of worker mode of transit by race/ethnicity. 

```{r, message=FALSE, warning=FALSE}

prop.table(xtabs(~ racehisp + mode, data = pums, subset = !is.na(mode), drop.unused.levels = TRUE), 1)

#Or if you saved the xtabs results in our object this also works
prop.table(xtabs.TR, 1)
```

A crosstab with column percentages can be quite useful for our analyses. When you present this sort of table in a report be sure to include total "N" or frequency counts for each cell.

**Question:** Now that we have some results, which racial groups are most likely to use a car, truck, or van to get to work?

##Run the Chi-Square

We will use the chi-square test in order to determine whether the difference between two groups is significant. We can use two approaches for getting a chi-square test for two groups using either the **summary()** function or the **chisq.tes()** function. 

```{r,message=FALSE, warning=FALSE}
#using our xtabs.TR object

summary(xtabs.TR)

chisq.test(xtabs(~racehisp + mode, data = pums, subset = !is.na(mode), drop.unused.levels = TRUE))
chisq.test(xtabs.TR)
```

Are the differences in modes to work between racial and ethnic groups statistically significant?

Save your script and email to me when finished **jamgreen@pdx.edu**

