---
title: "USP 634 Lab 3- Some R Basics"
author: "Jamaal Green"
date: "April 12, 2016"
output: pdf_document
---

##Introduction

I this lab we will be introduced to the beginning of our R workflow. This will include covering the basics of creating and deleting data, variable manipulation, and saving our results in various formats for later use.

*Load .RData file* .RData files are files that save whatever objects are in one's global environment at the time of saving. We do not need to cover all of the intricacies of differing environments, but do know that .RData files are a data type built for R and you do not need to have any additional packages to load them and the syntax is straightforward.

If you have not done so already please go to the ["Labs"]<https://sites.google.com/a/pdx.edu/usp634/labs> section of the course website and download the eb_pums_2000.RData file to your local folder and set your working directory.

```{r, echo=FALSE}
setwd("C:/Users/elmue/Desktop/Data1_labs")

load("C:\\Users\\elmue\\Desktop\\Data1_labs\\eb_pums_2000.RData")

#Note: you do not need to have the full file path if #you have properly set your wd(), you just need the #filename surrounded by quotation marks
```

After loading your data you can rename it by thinking of a name and assigning it the "data" dataframe:

> new name <- data

Now that we have our data loaded up, let's move onto the basics of creating and deleting data.

##Creating and Deleting Data

R supports vectors and matrices. You can create a scalar, vector, and matrix with the following command:
>    s <- 5 + 1

This creates a vector form the result 5+1. Now type "s" and hit return in order to get the answer.

>    v1 <- c(2,3,4)

This creates a vector of 2,3,4. Now type "v1" and hit return in order to see your vector.

>    v2 <- 1:10

This creates a vector of all integers between 1 and 10. Now type "v2" in order to view your vector. 

>    m <- matrix(c(1,2,3,11,12,13), nrow = 2, ncol = 3, byrow = T)

This creates a matrix of the above values. Now type "m" in order to see it. If you wish to learn more about the differences between matrices and data frames please see [here]<http://faculty.nps.edu/sebuttre/home/R/matrices.html#DataFrames>.

*Constructing Data Frames* Say you have collected some data with a few variables and you want to input it in R. We can use the data.frame() function

>    df <- data.frame(x = c(1,2,3,4,5), y = c(5,7,4,8,0), sex = c(1,2,2,1,2))

Type "df" in order to print out your data frame to your console. **Note: Data frames are the more "traditional rectangular data sets we are used to seeing and are the bread and butter data objects we will use in this class**

##Manipulate variables and values

*View Column or Row of Data Frame* View a variable column in a data frame with the "$" operator e.g.

> data$year

You can access rows of a data frame with this syntax: data[1:2,] (this will return the first two rows of your data frame not including your headers)

*Using Operators* In R you can use all the common mathematical and logical operators to create new variables.

The most common operators are <-, +, -,*,/,^,>,<.>=,<=

* <- is the operator for assigning a value to the variable a the left
* == means "equals" for conditional statements (conditional statements use an if expression such as "is such and such a condition is true")
* != means "does not equal"
* & means "and"
* | means "or"

*Creating Variables*

We have covered creating new variables using the basic [name of variable] <-. But how do we make a new variable for a data frame?

It is a straightforward process where we use the "$" operator and <- together, like so:

> data$new_variable <- (expression)

Here "data" is our dataframe and "new_variable" is the variable we want to define. 

* Mathematical operations on variables: We can create varibales basedon values of existing variables. For example, if you have a variable for total household income "inctot", create a new variable with total household income in thousands.

> data$inctot1k <- data$inctot/1000

* True/False from Existing Variables: Create a new variable "age65" in a new column at the end of the table, indiciating whether a person is older than 65 (using True/False)

> data$age65 <- data$age > 65

* Create new variables by recoding existing ones: We often need to recode existing variables into different ones for ease of analysis, visualization, or reporting. R has multiple ways and packages that can assist, but we will first cover the most basic approach to recoding. Assume we want to create a poverty status variable called "povstat" which takes the "1" when household income (hhinc) is less than the hypothetical poverty line of $12,000 or "0" if the household income is above the poverty line

> data$povstat <- ifelse(data$hhinc < 12000, 1,0)

* Factors and levels: Remember that R calls categorical variables "factors" and that factors can be nominal or ordinal (meaning they have an order). Say we want to create a factor variable from our new "povstat" variable where 1 = "below the poverty line" and 0 = "above poverty line"

> data$povfactor <- factor(data$povstat, levels = c(1,0), labels =c("below poverty line", "above poverty line"))

* Dropping columns: In order to remove a variable you can just type 

> data$variable_to_delete <- NULL

You can also use the <- NULL command to delete a dataframe, though the name will still be present in your environment. 

* Set an active data frame: you can tell R to set a data frame to an active data frame using the attach() function so you no longer have to use the $ operator. You would just have to type povstat instead of data$povstat. 

**BUT THIS IS NOT RECOMMENDED AS GENERAL PRACTICE! YOU WILL OFTEN HAVE MULTIPLE DATAFRAMES IN AN ENVIRONMENT AND ATTACHING CAN QUICKLY BECOME CONFUSING**

##Cleaning and Creating Data

First, we will create a new race/ethnicity variable. Remember that Census data include two race/ethnicity variables: race and Hispanic origin. We will use these categories to create the follow categories for our population: White non-Hispanic, lack non-Hispanic, Asian non-Hispanic, Hispanic, and Other non-Hispanic.

First we will make a quick two-way contingency table to take a look at our two variables.

```{r}

xtabs(~race + hispan, data = data)
```

To rceate the new variable, we must first understand the way the variables for race and ethnicity are organized and coded.

```{r}

#one-way table frequency tabulation
table(data$race)
table(data$hispan)

#to view the underlying codes 
table(as.integer(data$race))
table(as.integer(data$hispan))
```

Remember that factors (the name for categorical variables in R) can be ordered or unordered. As such, they have numeric values attached to them that R uses an identifier, so "White non-Hispanic" is coded as "1" and so on. 
*Creating New Variables and Recoding Part 2*

We have already covered the basics behind recoding. And as with all things R there are many ways to do it. In this following section we will use the **dplyr** pacakge to recode some variables. It is not required for you to only use it, but I find the dplyr syntax and logic to be fairly straightforward.

We will make what is called a nested conditional statement. This is a long statement for saying we are going to make a series of ifelse() statements to recode our variable.

**Note: You can use the label, such as "White" or "Black/Negro" or the underlying integer value of the factor such as 1 for "White" or 2 for "Black/Negro"**

```{r, message=FALSE, warning=FALSE, tidy=FALSE}

#install.packages("dplyr", repos = #'http://cran.us.r-project.org')

library(dplyr)

#we will create a new variable "racehisp" using #dplyr's mutate function
data <- data %>% mutate(racehisp = ifelse(hispan != "Not Hispanic", 6, 
       ifelse(race == "White" , 1, 
       ifelse(race == "Black", 2, 
       ifelse(race == "American Indian or Alaskan", 3, 
       ifelse(race == "Chinese" | race == "Japanese"| race == "Other Asian or Pacific", 4, 
       5))))))

#check your work
table(data$racehisp, data$race)
xtabs(~race + hispan, data = data)

#label new 
data$racehisp <- factor(data$racehisp, levels = 1:5,
labels = c("White non-Hisp", "Black non-Hisp", "Asian non-Hisp", "Hispanic", "Other"))

#check work again
table(data$racehisp)
```

**Recoding a continuous variable**

Now we will recode a continuous variable--personal tincome ("inctot"). We will recode inctot to an ordinal categorical variable (or a leveled factor variable in R parlance) by grouping categories of income.

*Look at the Data* 

Let's start by looking more closely at the orginal variable's coding and distribution by using the **summary()** and **hist()** functions.

```{r, message=FALSE, warning=FALSE}

summary(data$inctot)

hist(data$inctot)
```

Notice that the top values are repeated (999999). That code indicates "N/A" (see technical documentation at [www.ipums.org]<http://www.ipums.org>). There are also negative values indicating income losses. 

*Clean the Data*

We will clean this data by eliminating outliers and problematic cases. We will do this by creating a recode of the original vairable (Note: NEVER write over your original data).

> data$increc <- data$inctot

```{r}

#first we'll copy over our original variable to a
#new column
data$increc <- data$inctot

#Set Outliers to "NA". Values less than zero or
#999999 to NA

data$increc[data$inctot <= 0 | data$inctot >= 999999] <- NA


```

We will now recode the data into our new categories, in this case, quantiles. To do so, we shall use the **Hmisc** pacakge and its **cuts()** function to divide income into quintiles.

```{r}

install.packages("Hmisc",repos='http://cran.us.r-project.org' )

library(Hmisc)

#We will create a new variable called "inc5" made up #of 5 quintiles (6 slices of 5 categories) of our #cleaned up household income variable

data$inc5 <- cut2(data$increc, g = 6)

#check that it worked

summary(data$inctot)

summary(data$inc5)

#We will run a new two way table with our two new #variables "racehisp" and "inc5"

xtabs(~inc5 + racehisp, data = data)
```

Try comparing the results using these two methods (one using the categorical variable racehisp with an ordinal variable-- our quintile recode of income-- and the other using racehisp with the original interval-ratio version of income-- increc). What do they indicate?

##Saving Your Work

*Save R Output/Scripts*

In RStudio click File --> Save in order to save your scripts so you can reference them later on.

*Save Graphs*

Navigate to your "Plots" pane in RStudio and you can save individual graphs in varied formats such as a png or pdf file and modify its size.
